---
title: "Advanced features"
author: "magenta development team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_align: center
vignette: >
  %\VignetteIndexEntry{Advanced features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

In addition to the core malaria transmission dynamics, `magenta` implements a
number of auxiliary processes that are useful for specialised analyses. This
vignette describes three of them: non-malarial fevers, vector adaptation, and
explicit tracking of male and female sexual parasites.

```{r setup, eval = FALSE}
library(magenta)
```

## Non-malarial fevers

Non-malarial fevers (NMF) are modelled as a stochastic process that can trigger
healthcare-seeking behaviour even in the absence of parasitaemia. This is useful
when studying diagnostic specificity or the consequences of presumptive
antimalarial treatment.

The behaviour is controlled through `nmf_list_create()`, which bundles all
parameters required by the simulator:

```{r nmf, eval = FALSE}
nmf <- nmf_list_create(
  nmf_flag = TRUE,
  mean_nmf_frequency = c(148, 140, 142, 156, 179, 200),
  nmf_age_brackets = c(-0.1, 365, 730, 1095, 1460, 1825, 3650)
)
```

- `nmf_flag` toggles the feature on or off.
- `mean_nmf_frequency` gives the expected number of days between episodes for
  each age bracket.
- `nmf_age_brackets` sets the age cut-offs (in days) for those brackets.

The resulting list is supplied to `pipeline()` through the `nmf_list` argument.
During the simulation the loggers report the additional treatment demand caused
by these events, allowing downstream analysis to separate malaria-positive and
malaria-negative health facility visits.

## Vector adaptation

Vector adaptation captures changes in the vector population caused by
interventions such as insecticide-treated nets or indoor residual spraying. The
feature modulates the mosquito biting preference across the barcode loci that
influence transmissibility.

```{r vector-adaptation, eval = FALSE}
va <- vector_adaptation_list_create(
  vector_adaptation_loci = c(0, 1, 2),
  vector_adaptation_flag = TRUE
)
```

- `vector_adaptation_loci` identifies the loci that influence adaptation.
- `vector_adaptation_flag` activates or deactivates the feature.

As with the NMF configuration the returned list is passed to `pipeline()` via the
`vector_adaptation_list` argument. The impact on the force of infection can be
monitored through the standard logger outputs (`daily_bite_counters` and
`daily_infectious_bite_counters`).

## Male and female sexual parasites

`magenta` explicitly tracks the male and female contributions of sexual stage
parasites when generating new mosquito infections. This enables realistic
recombination and underpins the resistance and IBD modules.

While the underlying implementation lives in the C++ core (see `person.cpp` and
`mosquito.cpp`), several user-facing behaviours are worth noting:

- The infection summaries distinguish between male- and female-derived parasite
  contributions when constructing oocysts. This is reflected in the
  `populations_event_and_strains_List` element returned when `full_save = TRUE`.
- When simulating IBD the male and female barcode vectors are updated separately
  before being combined into the recombinant oocysts, preserving the pedigree
  information.
- The `daily_infectious_bite_counters` logger accounts for the number of
  infectious bites arising from male–female pairings rather than from individual
  parasites.

Users rarely need to interact with these components directly, but understanding
that they exist clarifies why certain summaries (for example `summary_within_ibd`)
exclude mono-infections: a single strain lacks the male–female pairing required
for recombination.

## Combining advanced features

All three features integrate with the standard `pipeline()` workflow:

```{r advanced-pipeline, eval = FALSE}
out <- pipeline(
  years = 5,
  N = 5000,
  EIR = 1.2,
  ft = 0.4,
  nmf_list = nmf,
  vector_adaptation_list = va,
  update_save = TRUE,
  summary_saves_only = TRUE,
  full_save = TRUE
)
```

Because they reuse the existing parameter list infrastructure they also work
when simulations are restarted from a saved state.

## Summary

- Non-malarial fevers can be activated via `nmf_list_create()` to capture
  treatment-seeking unrelated to malaria infections.
- Vector adaptation is controlled with `vector_adaptation_list_create()` and
  alters how parasite genotypes interact with the mosquito population.
- The simulator maintains separate male and female parasite contributions,
  enabling realistic recombination in both the standard and IBD modes.
- These features can be combined with other options such as `update_save` and
  `full_save` without additional boilerplate.
