---
title: "Drug components"
author: "magenta development team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_align: center
vignette: >
  %\VignetteIndexEntry{Drug components}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

Drug pressure is a key driver of the parasite genetics simulated by `magenta`.
The package provides helper functions for constructing rich drug definitions and
combining them into treatment policies. This vignette expands on the summary in
the resistance article and explains how to work with the drug helper functions.

```{r setup, eval = FALSE}
library(magenta)
```

## Building drug lists

The `pipeline()` function expects a `drug_list` element describing the drugs in
use. The `drug_list_create()` helper is the primary entry point:

```{r drug-list, eval = FALSE}
drug_list <- drug_list_create(
  drugs = list(magenta:::drug_create_dhappq()),
  partner_drug_ratios = 1,
  number_of_drugs = 1
)
```

Each entry in the list corresponds to a therapy. Inspecting the structure reveals
the parameters that `magenta` uses when simulating treatment outcomes.

```{r inspect-drug, eval = FALSE}
str(drug_list$drug_list[[1]])
```

The key components are:

- `barcode_loci`: integer indices indicating which loci of the parasite barcode
  are associated with the drug.
- `m_lpf`: probabilities of late parasitological failure (LPF) for every possible
  genotype, ordered lexicographically by the resistance loci. For six resistance
  loci this vector has length `2^6 = 64`.
- `prophylaxis_loci`: loci that determine the duration of prophylaxis provided by
  the treatment.
- `has_adherence` and `adherence`: switches that control whether imperfect
  adherence is modelled and the distribution of adherence outcomes.

All helper constructors ultimately call `drug_create()`, so custom therapies can
be defined by supplying bespoke arguments to that function. The package includes
convenience wrappers such as `drug_create_dhappq()` (dihydroartemisinin–piperaquine)
and `drug_create_al()` (artemether–lumefantrine) for the most commonly modelled
regimens.

## Treatment policies

Multiple drugs can be combined to simulate complex deployment strategies.
`partner_drug_ratios` controls the mixing between therapies, and
`mft_type = "cycling"` or `"mixing"` selects between sequential and simultaneous
use.

```{r multiple-drugs, eval = FALSE}
drug_list <- drug_list_create(
  drugs = list(
    magenta:::drug_create_al(),
    magenta:::drug_create_dhappq()
  ),
  partner_drug_ratios = c(0.5, 0.5),
  number_of_drugs = 2,
  mft_type = "mixing"
)
```

The returned object is passed unchanged to `pipeline()` via the `drug_list`
argument. During the simulation the active treatment is tracked in the loggers,
allowing the investigator to link observed resistance dynamics back to the
policy choices.

## Reference tables

The `drug_table` data set summarises the built-in therapies and can be used to
look up the mapping between barcode loci and genetic mechanisms:

```{r drug-table, eval = FALSE}
head(drug_table)
```

Each row represents a drug with columns describing the loci involved in
clearance, prophylaxis, partner drug protection, and the assumed efficacy. This
resource is a useful starting point when defining bespoke treatment policies.

## Summary

- Use `drug_list_create()` to compose the treatments available in a simulation.
- Helper constructors such as `drug_create_dhappq()` populate realistic defaults
  but custom calls to `drug_create()` are supported.
- The `drug_table` data object records the loci associated with each built-in
  therapy and can be extended for new scenarios.
