---
title: "Identity-by-descent simulations"
author: "magenta development team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_align: center
vignette: >
  %\VignetteIndexEntry{Identity-by-descent simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

`magenta` can run in a special identity-by-descent (IBD) mode where each barcode
locus stores a unique identifier rather than a binary allele. This enables the
model to track the pedigree of parasites through time and to calculate the
population-level (pIBD) and within-host (iIBD) measures described in the
accompanying manuscript. This vignette summarises how to configure IBD
simulations and interpret the resulting output.

```{r setup, eval = FALSE}
library(magenta)
```

## Allocating barcode space

IBD mode assigns consecutive integers to every parasite introduced during the
simulation. The `ibd_length` argument controls how many bits are allocated per
locus; the total number of unique identifiers is therefore `2^ibd_length`. To
avoid exhausting the identifier space, first run a dry setup with
`set_up_only = TRUE` and inspect the initialised population.

```{r setup-only, eval = FALSE}
out <- pipeline(
  EIR = 1,
  years = 10,
  N = 1000,
  ibd_length = 20,
  num_loci = 24,
  set_up_only = TRUE,
  update_save = TRUE,
  update_length = 30,
  human_update_save = TRUE,
  summary_saves_only = TRUE,
  full_save = TRUE,
  seed = 123456789L
)
```

The `population_List` and `populations_event_and_strains_List` entries describe
the seeded human population. Each infected host contains a list of IBD barcodes
that can be converted back to integers with the unexported
`convert_ibd_barcode()` helper.

```{r inspect-barcode, eval = FALSE}
strains <- out$populations_event_and_strains_List$Strain_barcode_vectors
infected <- which(out$population_List$Infection_States %in% 1:4)
magenta:::convert_ibd_barcode(strains[[infected[1]]][[1]], nl = 24)
```

If the largest identifier in the initial state is close to `2^ibd_length` then a
larger `ibd_length` is required before running long simulations or adding heavy
importation.

## Running and checking simulations

A regular simulation in IBD mode is started by passing the same arguments but
leaving `set_up_only` at its default of `FALSE`.

```{r ibd-run, eval = FALSE}
out <- pipeline(
  EIR = 1,
  years = 10,
  N = 1000,
  ibd_length = 20,
  num_loci = 24,
  update_save = TRUE,
  update_length = 30,
  human_update_save = TRUE,
  summary_saves_only = TRUE,
  full_save = TRUE,
  seed = 987654321L
)
```

At the end of the run the final snapshot in the returned list contains the full
state of the system. Inspecting the `Strain_barcode_vectors` confirms that the
identifier space was sufficient:

```{r ibd-final, eval = FALSE}
final <- out[[length(out)]]
strains <- final$populations_event_and_strains_List$Strain_barcode_vectors
infected <- which(final$population_List$Infection_States %in% 1:4)
converted <- lapply(strains[infected], function(x) lapply(x, magenta:::convert_ibd_barcode, nl = 24))
max(unlist(converted))
```

## Understanding IBD summaries

When IBD mode is active the monthly summaries include two additional entries:

- `summary_ibd` stores the population-level pIBD stratified by infection state
  and age groups.
- `summary_within_ibd` stores the within-host iIBD statistics, calculated only
  for hosts carrying more than one strain.

Both summaries share the same structure as the barcode summaries in the standard
mode and can be converted to tidy tibbles for plotting or further analysis.

## Summary

- Set `ibd_length` high enough to cover the expected number of unique parasites;
  use `set_up_only = TRUE` to check the initial allocation.
- Identity-by-descent simulations work seamlessly with `update_save`,
  `summary_saves_only`, and `full_save`.
- Monthly summaries include both pIBD and iIBD measures, enabling direct
  comparison with the outputs reported in the manuscript.
