---
title: "Loggers and model state"
author: "magenta development team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_align: center
vignette: >
  %\VignetteIndexEntry{Loggers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

`magenta` summarises each simulation through a set of *loggers*. These are
lists of scalar and tabular outputs that track high-level properties of the
human and mosquito populations without needing to retain every individual in
memory. This vignette describes the key loggers that are returned by
`pipeline()` and how to control the different saving options.

Throughout we assume that the package has been attached and that simulations are
run with sensible settings. The code examples below set `eval = FALSE` so that
they can be inspected without triggering a long simulation during vignette
building.

```{r setup, eval = FALSE}
library(magenta)
```

## Inspecting the default loggers

By default `pipeline()` returns a list with two components: the Rcpp pointer to
the simulation universe and a `Loggers` list containing the final summaries.
Running a short simulation and examining the structure illustrates what is
available.

```{r default-loggers, eval = FALSE}
out <- pipeline(years = 5, N = 1000, EIR = 1, seed = 123)
str(out$Loggers)
```

Some commonly used elements include:

- `Successful_Treatments`, `Unsuccesful_Treatments_LPF`, and `Not_Treated`
  summarising the most recent treatment events.
- `daily_bite_counters` and `daily_infectious_bite_counters` recording vector
  activity.
- `S`, `D`, `A`, `U`, `T`, and `P`, which contain the proportion of the human
  population in each infection state at the end of the run.

Because the loggers are standard R lists they can be converted into data frames
or other structures for downstream analysis.

## Saving intermediate updates

Longer simulations often require insight into the temporal evolution of the
system. Setting `update_save = TRUE` instructs `pipeline()` to stop after every
`update_length` days and append the requested summaries to the output list.

```{r update-loggers, eval = FALSE}
out <- pipeline(
  years = 5,
  N = 1000,
  EIR = 1,
  update_save = TRUE,
  update_length = 30,
  seed = 456
)
length(out)          # monthly snapshots plus the final state
names(out[[1]])      # each snapshot contains the familiar Loggers list
```

Two additional switches refine the behaviour:

- `human_update_save = TRUE` saves the full human population at every update
  interval, allowing complex genetic summaries to be recomputed after the run.
- `summary_saves_only = TRUE` keeps only the logger summaries for each update,
  avoiding the large memory footprint associated with full population saves.

The Loggers vignette complements the *Advanced features* article, which
provides more practical tips on choosing update intervals for large scenarios.

## Persisting and restoring model state

`magenta` supports saving the full universe state so that simulations can be
paused and resumed. Setting `full_save = TRUE` (optionally with
`human_only_full_save = TRUE`) adds several additional entries to the list
returned by `pipeline()`, including
`populations_event_and_strains_List` and `population_List`. These contain the
R objects required by `Simulation_Saved_Init_cpp()` to restart a run.

```{r full-save, eval = FALSE}
out <- pipeline(
  years = 2,
  N = 500,
  EIR = 1.5,
  full_save = TRUE,
  summary_saves_only = FALSE,
  seed = 789
)
```

The combination of the loggers, the saved human population, and the parameter
lists (`model_param_list`, `drug_list`, `nmf_list`, and so on) is sufficient to
serialise a simulation to disk. See the advanced features vignette for worked
examples on saving these components with `qs`, `saveRDS()`, or other
serialisation helpers.

## Logger summaries for genetics

When `summary_saves_only = TRUE` the simulation records monthly summaries of the
parasite genetics:

- mean multiplicity of infection (COI) by age and infection state;
- proportion of polygenomic infections and proportion of unique barcodes;
- allele frequencies across the modelled barcode;
- optional summaries for identity-by-descent (pIBD and iIBD) when simulating in
  IBD mode.

The summaries live within the `summary_stats` element of the loggers and are
organised as nested lists keyed by the epidemiological groups. These objects are
designed to be tidy-friendly; each is a data frame or matrix with columns for
the bin definitions and summary statistics. The helper functions in
`plotting.R` demonstrate how to convert the summaries into plots.

## Summary

- `pipeline()` always returns a `Loggers` list containing the final simulation
  summaries.
- `update_save`, `human_update_save`, and `summary_saves_only` control whether
  intermediate snapshots or full human populations are retained.
- `full_save` and `human_only_full_save` add the state required to resume a
  simulation from an arbitrary checkpoint.
- The logger summaries include rich information on parasite genetics that feeds
  into downstream analysis and plotting utilities.
